---
title: "Diffusion capacity (DLCO / TLCO): reference, Hb correction, interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diffusion capacity (DLCO / TLCO): reference, Hb correction, interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(pft)
```

Diffusion measurements (DLCO in traditional units, TLCO in SI)
assess gas-exchange capacity at the alveolar-capillary membrane and
are typically added to spirometry and volumes when the differential
includes interstitial lung disease, pulmonary vascular disease,
emphysematous parenchymal loss, or anemia. The sections below cover
reference values, hemoglobin correction, the six Hughes & Pride
clinical categories, and the alveolar-volume distinction between
parenchymal and extra-parenchymal restriction.

# 1. Reference values

`pft_diffusion()` implements the GLI 2017 standard (Stanojevic et al.
ERJ 2017, with the 2020 author correction applied) for adults and
children aged 4-85 years. By default it emits **traditional units**
(DLCO, KCO, VA in mL/min/mmHg, mL/min/mmHg/L, and L respectively);
`SI.units = TRUE` switches to SI units (TLCO and KCO in mmol/min/kPa
and mmol/min/kPa/L). VA is the same column either way.

```{r}
patient <- data.frame(
  sex = "M", age = 45, height = 178,
  dlco_measured   = 22.0,
  va_measured     = 5.8,
  kco_tr_measured = 3.79
)
out <- pft_diffusion(patient)
out[, grep("dlco|va|kco", colnames(out), value = TRUE)]
```

The output carries the same per-measure `_pred` / `_lln` / `_uln` /
`_zscore` / `_pctpred` shape as `pft_spirometry()` and
`pft_volumes()`.

# 2. Hemoglobin correction

DLCO measured against the standard reference Hb may misrepresent
patients who are anemic (DLCO under-reads) or polycythemic (DLCO
over-reads). `pft_dlco_hb_correct()` applies the Cotes 1972 formula
to express the measured DLCO at the standard reference Hb:

\[
\text{DLCO}_\text{adj} =
  \text{DLCO} \cdot \frac{1.7 \cdot \text{Hb}_\text{ref}}{\text{Hb} + 0.7 \cdot \text{Hb}_\text{ref}}.
\]

The reference Hb is age- and sex-dependent: 146 g/L for males aged
>= 15, 134 g/L for females and for children < 15 of either sex
(Cotes 1972 / Stanojevic 2017 Table 5).

```{r}
# Anemic adult male: corrected DLCO is higher than measured.
pft_dlco_hb_correct(dlco = 20.0, hemoglobin = 110, sex = "M", age = 45)

# Polycythemic adult male: corrected is lower than measured.
pft_dlco_hb_correct(dlco = 25.0, hemoglobin = 180, sex = "M", age = 45)
```

The function also accepts Hb in g/dL (auto-detected when values are <
30) and converts internally. **Apply the correction before
computing z-scores** when comparing across patients whose Hb varies;
the GLI 2017 reference values assume Hb is at the sex-/age-specific
standard.

# 3. Clinical sub-pattern (Hughes & Pride 2012)

`pft_diffusion_interpret()` classifies a diffusion result into one of
six clinical categories per the Hughes & Pride 2012 framework
(adopted by the Stanojevic 2017 task force). The classifier uses
z-scores only, so it works identically on traditional and SI
columns:

```{r}
mixed_cohort <- data.frame(
  dlco_zscore   = c(-0.5, -2.0, -2.5, -2.5, -2.0,  0.0),
  va_zscore     = c(-0.5, -0.5, -2.0, -2.5, -0.5,  0.0),
  kco_tr_zscore = c(-0.5, -2.0,  0.0, -2.5,  0.5,  2.0)
)
pft_diffusion_interpret(mixed_cohort)
```

The decision tree (Stanojevic 2017 / Hughes & Pride 2012):

| Category               | DLCO | VA | KCO | Typical context |
|------------------------|------|----|----|------------------|
| **Normal**             | OK   | OK | OK | -- |
| **Parenchymal**        | low  | OK | low | ILD, emphysema, anemia, COHb |
| **Volume loss**        | low  | low | OK / high | Chest wall, neuromuscular, post-lobectomy, atelectasis |
| **Mixed**              | low  | low | low | Combined volume loss + gas-exchange |
| **Vascular (suggested)** | low | OK | low or high | PE, PAH (capillary bed impaired but alveolar volume preserved) |
| **Elevated KCO**       | OK   | -- | high | Hyperventilation, polycythemia, recovery from anemia |
| **Other**              | other combinations | | | -- |

In `pft_interpret()` the classifier runs automatically whenever the
diffusion z-score columns are present, so the
`diffusion_category` column is attached for free in the standard
workflow:

```{r}
patient2 <- data.frame(
  sex = "F", age = 60, height = 165, race = "Caucasian",
  fev1_measured    = 1.6, fvc_measured    = 1.9,
  fev1fvc_measured = 0.84, tlc_measured   = 4.0,
  dlco_measured   = 10.0, va_measured    = 3.5,
  kco_tr_measured = 2.86
)
r <- pft_interpret(patient2)
r[, c("ats_classification", "diffusion_category")]
```

# 4. VA: distinguishing parenchymal from extra-parenchymal restriction

Alveolar volume (VA) carries diagnostic weight on its own. In
restriction, VA tracks total alveolar gas exchange capacity:

* **Low VA, low KCO** -> parenchymal restriction (the alveoli that
  *are* working are also impaired): suspect interstitial lung
  disease, advanced fibrosis, or a chronic process that has
  degraded both volume and gas exchange.
* **Low VA, normal-or-high KCO** -> extra-parenchymal restriction
  (fewer alveoli, but each functions normally per unit volume):
  suspect chest-wall or neuromuscular causes, post-surgical loss of
  lung tissue, or atelectasis.

The Hughes & Pride classifier encodes this distinction: low-DLCO +
low-VA + low-KCO is **Mixed**, low-DLCO + low-VA + normal-KCO is
**Volume loss**.

# 5. Cohort-level diffusion summaries

`pft_cohort_summary()` includes a `diffusion` component when the
input data frame carries a `diffusion_category` column (e.g. a
`pft_interpret()` result):

```{r}
cohort <- data.frame(
  sex    = c("M","F","M","F","M","F"),
  age    = c(45,60,30,55,70,28),
  height = c(178,165,175,160,170,180),
  race   = "Caucasian",
  fev1_measured    = c(2.5, 1.8, 4.0, 1.5, 2.2, 3.8),
  fvc_measured     = c(3.8, 2.4, 5.2, 2.5, 3.5, 5.0),
  tlc_measured     = c(6.0, 4.5, 6.8, 4.0, 6.5, 7.0),
  dlco_measured    = c(20.0, 12.5, 28.0, 10.0, 18.0, 25.0),
  va_measured      = c(5.8, 4.0, 6.5, 3.5, 5.5, 6.0),
  kco_tr_measured  = c(3.5, 3.2, 4.3, 2.9, 3.3, 4.0)
)
r <- pft_interpret(cohort)
s <- pft_cohort_summary(r, by = "sex")
s$diffusion
```

# When NOT to order a diffusion test

Diffusion measurement adds time and cost to a PFT session. Typical
indications per the standard:

* Suspected interstitial lung disease (any pattern with restriction).
* Suspected emphysema or chronic parenchymal damage (obstructed
  pattern with no clear hyperinflation).
* Pulmonary vascular disease evaluation.
* Pre-operative risk stratification for lung resection.
* Drug-toxicity monitoring (bleomycin, amiodarone, methotrexate).

Diffusion is *not* generally recommended for routine spirometry
follow-up of stable COPD or asthma; the marginal information
typically isn't worth the test burden. The standard's recommendation
is "order when the result would change management".

# See also

* `vignette("interpretation-guide")` -- pattern decision tree and
  severity bands.
* `vignette("longitudinal-analysis")` -- serial DLCO change and
  decline.
* `?pft_diffusion`, `?pft_diffusion_interpret`,
  `?pft_dlco_hb_correct` for the function references.
