---
title: "Input data format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input data format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
library(pft)
```

The reference functions in `pft` (`pft_spirometry()`, `pft_volumes()`,
`pft_diffusion()`, and the one-call workflow `pft_interpret()`) all take
a data frame and append reference columns to it. This article documents
exactly which columns each function expects, which are optional, and
how to point pft at columns whose names don't match the canonical ones.

# Required vs. optional columns

`pft_required_columns()` is the authoritative source for the input
contract. It returns a list with two character vectors per function:

```{r}
pft_required_columns("pft_spirometry", year = 2012)
pft_required_columns("pft_spirometry", year = 2022)
pft_required_columns("pft_volumes")
pft_required_columns("pft_diffusion", SI.units = FALSE)
pft_required_columns("pft_interpret", year = 2012)
```

**`required`** -- columns that must be present. Missing required
columns raise an error.

**`optional_measured`** -- columns whose presence unlocks z-score and
percent-predicted outputs for that measure. Missing these is silent;
the function simply emits fewer output columns.

**`optional_bdr`** -- (`pft_interpret` only) `<measure>_pre` and
`<measure>_post` columns that, when present, unlock bronchodilator
response calculations.

# Units and types

The canonical types and units are:

| Column         | Type      | Allowed values / units                                    |
|----------------|-----------|-----------------------------------------------------------|
| `sex`          | character | `"M"` or `"F"`. Common variants (`"male"`, `"Female"`, `"m"`, ...) are auto-normalised with a warning. |
| `age`          | numeric   | Years (decimal allowed).                                  |
| `height`       | numeric   | Centimetres.                                              |
| `race`         | character | One of `"Caucasian"`, `"AfrAm"`, `"NEAsia"`, `"SEAsia"`, `"Other/mixed"`. Common synonyms are auto-normalised. (GLI 2012 only.) |
| `fev1_measured`, `fvc_measured`, etc. (spirometry) | numeric | Litres for volumes; L/s for flows; dimensionless for FEV1/FVC. |
| `frc_measured`, `tlc_measured`, ... (volumes)      | numeric | Litres. `rv_tlc_measured` is dimensionless. |
| `dlco_measured` / `tlco_measured` / `kco_*` / `va_measured` (diffusion) | numeric | Diffusion measures in the unit system requested via `SI.units`. |
| `<measure>_pre`, `<measure>_post` (BDR)            | numeric | Same units as `<measure>_measured`. |

See the [Glossary](glossary.html) for definitions of every measure
abbreviation.

# Auto-detecting common unit mistakes

The two most common data-import bugs that yield silently wrong
reference values are height in inches (instead of cm) and measured
volumes in millilitres (instead of L). `pft_normalize_units()`
detects both heuristically and converts in place, emitting a single
consolidated warning:

```{r}
suppressWarnings(
  pft_normalize_units(data.frame(
    sex = "M", age = 45, height = 70,   # 70 inches!
    race = "Caucasian",
    fev1_measured = 2500                # millilitres!
  ))
)
```

Detection is conservative -- the heuristic only fires on
unambiguous input. Pipe it before the rest of the workflow when
the data source is uncertain:

```r
df |> pft_normalize_units() |> pft_interpret()
```

# Using non-canonical column names

If your data frame uses different column names for the demographics,
all reference functions accept `sex`, `age`, `height`, and (where
applicable) `race` overrides. Three forms are accepted: a bare column
name (the usual tidyverse style), a string, or an injection from a
variable:

```{r}
patient <- data.frame(
  Sex      = "M",
  Age_y    = 45,
  Ht_cm    = 178,
  Ancestry = "Caucasian"
)

# Bare names (tidyverse-style)
pft_spirometry(patient,
               sex    = Sex,
               age    = Age_y,
               height = Ht_cm,
               race   = Ancestry)

# Strings -- equivalent
pft_spirometry(patient,
               sex    = "Sex",
               age    = "Age_y",
               height = "Ht_cm",
               race   = "Ancestry")

# Injection from a variable, e.g. driven by a config
sex_col <- "Sex"
pft_spirometry(patient, sex = !!sex_col,
               age    = Age_y,
               height = Ht_cm,
               race   = Ancestry)
```

The user's original column names are **preserved** in the output (no
renaming to canonical). Sex and race values are normalised in place,
so e.g. `"male"` becomes `"M"` in the original `Sex` column.

The `_measured`, `_pre`, and `_post` columns are not overridable --
they are looked up by name. If your data uses different names, rename
them before calling `pft`:

```{r, eval = FALSE}
library(dplyr)
patient %>%
  rename(fev1_measured = FEV1_L,
         fvc_measured  = FVC_L) %>%
  pft_spirometry(year = 2022)
```

# Pre-flight QC

Before running the reference functions on a fresh cohort, call
`pft_validate()` to get a row-level breakdown of input issues
(missing demographics, unrecognised sex/race values, implausible
heights or ages):

```{r, eval = FALSE}
pft_validate(my_cohort)
```

`pft_validate()` does not modify the data; it just reports.

# Common errors

**`required column(s) missing from input: 'race'`** -- you called
`pft_spirometry(d, year = 2012)` without a `race` column. Either
supply one (canonical levels listed in
`pft_required_columns("pft_spirometry", year = 2012)$required`) or
call `pft_spirometry(d, year = 2022)` for the race-neutral equations.

**Warning: `pft input normalization: ...`** -- one or more sex or
race values were normalised (e.g. `"male"` → `"M"`) or set to NA
because they didn't match any known canonical value. The warning is
consolidated to one message per call. Check the cohort with
`pft_validate()` to see which rows are affected.

**`year = 2012` running but all outputs NA** -- check whether the
`race` column contains values outside the five GLI 2012 categories.
Unknown values are set to NA, which propagates through the
reference equations.
