---
title: "Interpreting pft results: from z-score to clinical action"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting pft results: from z-score to clinical action}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(pft)
```

Where the Getting-started vignette covers *how* to compute reference
values and z-scores, this one covers *what to do with them*: the
severity bands, the pattern decision tree, the differences between
the 2022 Stanojevic and 2005 Pellegrino standards, and worked case
studies mapping common z-score / pattern combinations to the next
clinical step.

# Severity bands

`pft_severity()` translates a z-score into one of four bands per the
Stanojevic 2022 standard. The cut-points come straight from the
paper's interpretation table:

```{r}
data.frame(
  band     = c("normal", "mild", "moderate", "severe"),
  z_lower  = c(-1.645, -2.5, -4,    -Inf),
  z_upper  = c( Inf,   -1.645, -2.5, -4)
)
```

A vectorised call:

```{r}
pft_severity(c(0.2, -1.7, -3.0, -5.0))
```

The 2005 Pellegrino bands grade *percent predicted* of FEV1 rather
than z-score and have five tiers (mild, moderate, moderately-severe,
severe, very-severe). They are appropriate when reproducing legacy
reports or when matching a clinic's existing severity-grading
convention; use `pft_severity_2005()`:

```{r}
pft_severity_2005(c(85, 65, 55, 40, 30))
```

The same `standard = c("2022", "2005")` argument flows through
`pft_classify()`, `pft_compare()`, and `pft_interpret()` so a whole
report can be re-rendered against either standard without changing
input data.

# Pattern decision tree

`pft_classify()` assigns one of five interpretive patterns per
Stanojevic 2022 Figure 8 / Table 5:

* **Normal** -- FEV1/FVC, FVC, and FEV1 all >= LLN.
* **Obstructed** -- FEV1/FVC < LLN.
* **Restricted** -- FEV1/FVC >= LLN, FVC < LLN, *and* TLC < LLN.
* **Mixed** -- FEV1/FVC < LLN *and* TLC < LLN.
* **Non-specific** -- FEV1/FVC >= LLN, FVC < LLN, TLC >= LLN. The
  spirometry-only version of this pattern (TLC unavailable) is
  PRISm, surfaced by `pft_prism()`.

When TLC is missing, the classifier falls back to the
spirometry-only branches in Table 5 (Normal, Obstructed,
Non-specific / PRISm); Restricted and Mixed require TLC.

```{r}
case <- data.frame(
  fev1    = c(2.5, 2.5, 1.5, 1.5,  3.5),
  fev1_lln= c(3.0, 3.0, 2.5, 2.5,  3.0),
  fvc     = c(3.8, 3.8, 2.2, 2.2,  4.5),
  fvc_lln = c(3.5, 3.5, 2.5, 2.5,  4.0),
  fev1fvc = c(0.66, 0.66, 0.68, 0.80, 0.78),
  fev1fvc_lln = 0.70,
  tlc     = c(6.0, 5.5, 4.0, 4.0,  6.5),
  tlc_lln = c(5.5, 5.5, 5.5, 5.5,  5.5)
)
pft_classify(case)[, c("ats_classification")]
```

Reading row by row:

1. FEV1/FVC < LLN, TLC normal -> **Obstructed**.
2. FEV1/FVC < LLN *and* TLC < LLN -> **Mixed**.
3. FEV1/FVC normal, FVC < LLN, TLC < LLN -> **Restricted**.
4. FEV1/FVC normal, FVC < LLN, TLC normal -> **Non-specific**.
5. Everything >= LLN -> **Normal**.

# When to use 2022 vs 2005

The two standards differ in three ways:

| Aspect                | 2022 (Stanojevic) | 2005 (Pellegrino) |
|-----------------------|-------------------|-------------------|
| Severity input        | z-score           | % predicted (FEV1) |
| Bronchodilator response | > 10 % predicted | >= 12 % AND >= 200 mL |
| Pattern flowchart     | Fig 8 / Table 5    | Fig 2             |

The 2022 standard is the recommended default and is what
`pft_interpret()` applies by default. Use the 2005 path when:

* Reproducing a historical report or matching an EMR template that
  was built against the older flowchart.
* Performing a reclassification audit -- running the same cohort
  through both standards highlights patients whose label changed
  with the migration. `pft_compare()` produces the side-by-side
  result; `pft_cohort_summary()` then aggregates it.

```{r, eval = FALSE}
cohort_compare <- pft_compare(my_cohort)         # runs both
summary(cohort_compare)                          # cohort report
pft_cohort_summary(cohort_compare)$reclassification
```

# Case studies

## Case 1: moderate obstruction (typical COPD work-up)

```{r}
copd <- data.frame(
  sex = "M", age = 68, height = 175, race = "Caucasian",
  fev1_measured    = 1.6,
  fvc_measured     = 3.0,
  fev1fvc_measured = 1.6 / 3.0,
  tlc_measured     = 6.8
)
r <- pft_interpret(copd)
r[, c("ats_classification", "fev1_severity", "fev1_zscore",
       "fev1_pctpred")]
```

The pattern is **Obstructed** with **moderate** severity. GOLD
staging (FEV1 % predicted) classifies this as **GOLD 2**:

```{r}
pft_gold(r$fev1_pctpred, fev1fvc = r$fev1fvc_measured)
```

## Case 2: restriction with normal diffusion vs reduced diffusion

```{r}
extra_parenchymal <- data.frame(
  sex = "F", age = 55, height = 160, race = "Caucasian",
  fev1_measured    = 1.2, fvc_measured     = 1.5,
  fev1fvc_measured = 0.80, tlc_measured    = 3.8,
  rv_tlc_measured  = 0.30, dlco_measured   = 22.0,
  va_measured      = 4.6,  kco_tr_measured = 4.5
)
r <- pft_interpret(extra_parenchymal)
r[, c("ats_classification", "diffusion_category",
       "volume_subpattern")]
```

Restricted lungs with **Volume loss** diffusion pattern (low DLCO,
low VA, preserved KCO) and a **Simple restriction** sub-pattern
point toward extra-parenchymal restriction (chest wall, neuromuscular,
post-lobectomy). Contrast with a parenchymal restriction case where
KCO would also be low.

## Case 3: PRISm without TLC

When TLC isn't available, `pft_prism()` flags the spirometry-only
non-specific picture: low FEV1, low FVC, preserved ratio.

```{r}
no_tlc <- data.frame(
  sex = "M", age = 50, height = 175, race = "Caucasian",
  fev1_measured    = 2.2, fvc_measured     = 2.8,
  fev1fvc_measured = 0.79
)
r <- pft_interpret(no_tlc)
r[, c("ats_classification", "prism")]
```

The patient screens positive for **PRISm**. Stanojevic 2022 notes
PRISm is associated with elevated all-cause mortality and progression
to chronic respiratory disease; follow-up is recommended even though
the pattern itself isn't disease-specific.

# See also

* `vignette("longitudinal-analysis")` -- decline, conditional change,
  FEV1Q.
* `vignette("diffusion-capacity")` -- DLCO interpretation, Hb
  correction, Hughes & Pride categories.
* `vignette("input-format")` -- input contract and `pft_validate()`
  worked examples.
