---
title: "Longitudinal PFT analysis: change, decline, FEV1Q"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal PFT analysis: change, decline, FEV1Q}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
library(pft)
set.seed(7)
```

Serial PFTs are how pulmonary disease is actually managed: a single
result is rarely enough; what matters is whether the patient is
stable, declining, or recovering. `pft` provides three complementary
tools for this:

| Tool | Use case | Inputs |
|------|----------|--------|
| `pft_change()` | Two-point conditional change z-score (Stanojevic 2022 Box 2) | Two FEV1 z-scores + age + elapsed time |
| `pft_decline()` | Per-patient slope fit from 3+ time points | Long-form data: patient ID, measure, time |
| `pft_decline_grouped()` | Cohort-level slope per group (disease, smoking status) | Same + a grouping covariate |
| `pft_fev1q()` | Adult mortality index, alternative to CCS in adults | FEV1 (litres) + sex |

This vignette walks through each.

# 1. Two-point conditional change score: `pft_change()`

`pft_change()` evaluates whether the change between *two* FEV1
z-scores is larger than expected from within-subject variability and
regression to the mean. The formula (Box 2 of Stanojevic 2022):

\[
\text{CCS} = \frac{z_2 - r \cdot z_1}{\sqrt{1 - r^2}}, \quad
r = 0.642 - 0.04 \cdot \text{time}_\text{years} + 0.020 \cdot \text{age}_\text{years}.
\]

`|CCS| > 1.96` is the two-sided 95 % significance threshold:

```{r}
# Box 2 worked example: 14-year-old male, FEV1 z dropped from -0.78
# to -1.60 over 3 months.
pft_change(z1 = -0.78, z2 = -1.60, age_t1 = 14, time_years = 0.25)
```

The same drop spread over four years is not significant -- the
autocorrelation falls, so the same z-score change is more readily
explained by noise:

```{r}
pft_change(z1 = -0.78, z2 = -1.60, age_t1 = 14, time_years = 4)
```

**Scope.** The formula was derived in a pediatric / young-adult
cohort; the standard notes it has "yet to be validated, extended to
adults" but allows it as "a reasonable tool to facilitate
interpretation". For adults the standard recommends FEV1Q instead
(Box 3); see section 4 below.

# 2. Per-patient slope: `pft_decline()`

With three or more serial measurements, fit a slope. `pft_decline()`
takes long-form data (one row per (patient, timepoint)) and returns
one row per patient with the slope, standard error, 95 % CI, and the
observation count it was fit on.

```{r}
serial <- data.frame(
  patient_id = rep(1:3, each = 5),
  year       = rep(2018:2022, times = 3),
  fev1_zscore = c(
    -0.5, -0.4, -0.6, -0.5, -0.5,     # stable
    -0.5, -0.8, -1.2, -1.7, -2.1,     # declining
     0.2, -0.5, -1.4, -2.3, -3.2      # rapid decline
  )
)
pft_decline(serial, by = patient_id, measure = "fev1_zscore",
            time = year, flag_threshold = 0.25)
```

`flag_threshold` sets a clinically-meaningful cut-off. The unit is
"change in `measure` per unit of `time`" -- here, z-score per year.
The package deliberately does not bundle disease-specific thresholds
because they vary: CF Foundation uses 10 % FEV1/year, ISHLT 10 %
sustained FEV1 drop for BOS, OMERACT/ATS 5 % FVC over 6 months for
IPF, and so on. Pick the threshold appropriate to your clinical
context.

## OLS vs mixed-effects

`pft_decline(model = "ols")` (default) fits a per-patient `lm()`.
`pft_decline(model = "mixed")` fits a single `lme4::lmer()` with
random intercepts and slopes per patient. Mixed-effects partial-pools
toward the cohort-wide slope, which stabilises estimates for
patients with few observations at the cost of pulling outliers
toward the mean:

```{r, eval = requireNamespace("lme4", quietly = TRUE)}
pft_decline(serial, by = patient_id, measure = "fev1_zscore",
            time = year, model = "mixed")
```

Use OLS for per-patient inference (the SEs and CIs are valid for the
individual patient). Use mixed when individual fits are too noisy and
a cohort-anchored estimate is more useful than a per-patient one.

## Date / POSIXct support

`time` can be a numeric column, a Date column, or POSIXct. Date /
POSIXct are converted internally to years since the earliest
observed time. The same input is accepted by `pft_plot(type =
"trajectory")`:

```{r}
serial$visit_date <- as.Date(rep(c("2018-06-15","2019-07-20",
                                    "2020-09-01","2021-10-12",
                                    "2022-11-30"), 3))
pft_decline(serial, by = patient_id, measure = "fev1_zscore",
            time = visit_date)
```

# 3. Group-stratified decline: `pft_decline_grouped()`

When you want to know the *cohort-level* decline per group rather
than each patient's individual rate, `pft_decline_grouped()` fits a
single mixed-effects model with a `time:group` interaction and
returns one row per group level:

```{r, eval = requireNamespace("lme4", quietly = TRUE)}
n_patients <- 30
n_visits   <- 5
groups     <- c("never", "ex", "current")
slope_per_group <- c(never = -0.05, ex = -0.10, current = -0.30)

patient_group <- rep(groups, each = n_patients / length(groups))
cohort <- data.frame(
  patient_id = rep(seq_len(n_patients), each = n_visits),
  year       = rep(0:(n_visits - 1), times = n_patients),
  smoking    = rep(patient_group, each = n_visits)
)
cohort$fev1_zscore <- slope_per_group[cohort$smoking] * cohort$year +
                       rnorm(nrow(cohort), sd = 0.2)

pft_decline_grouped(cohort,
                     by = patient_id, measure = "fev1_zscore",
                     time = year, group = smoking)
```

This answers "is decline faster in current smokers?" with a single
model rather than averaging per-patient slopes. The SE / CI come
from the model's variance-covariance matrix.

# 4. Adult FEV1Q: `pft_fev1q()`

For *adults*, the 2022 standard recommends FEV1Q over the conditional
change score: the ratio of FEV1 to a sex-specific 1st-percentile
"bottom line" required for survival (Box 3). FEV1Q is not a change
score -- it characterises the *absolute* impairment relative to a
survivable lower limit, in a way that's interpretable across
patients and over time.

```{r}
# Box 3 worked example: 70-year-old female with FEV1 = 0.9 L.
pft_fev1q(fev1 = 0.9, sex = "F", age = 70)
```

Values closer to 1 indicate higher mortality risk; under normal
ageing one unit of FEV1Q is lost approximately every 18 years (10
years in smokers and the elderly). The function refuses age < 18 by
returning `NA_real_` per the paper's "not appropriate for children
and adolescents" caveat.

# 5. Plotting trajectories

`pft_plot(type = "trajectory")` overlays z-scores over time with the
2022 severity bands behind. It works on long-form data with a
`time` column (numeric, Date, or POSIXct) and an optional
`patient_id` column for faceting:

```{r}
pft_plot(serial, type = "trajectory",
         time = visit_date, patient_id = patient_id)
```

# See also

* `vignette("interpretation-guide")` -- severity bands, pattern
  decision tree, 2022 vs 2005.
* `vignette("diffusion-capacity")` -- DLCO interpretation and
  Hb correction.
* `?pft_change`, `?pft_decline`, `?pft_decline_grouped`,
  `?pft_fev1q` for the function references.
