---
title: "Longitudinal PFT analysis: change and FEV1Q"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal PFT analysis: change and FEV1Q}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
library(pft)
set.seed(7)
```

PFTs are often interpreted serially rather than from a single result.
`pft` provides two PFT-specific tools defined by the Stanojevic 2022
standard that operate on serial measurements:

| Tool | Use case | Inputs |
|------|----------|--------|
| `pft_change()` | Two-point conditional change z-score (Stanojevic 2022 Box 2) | Two FEV1 z-scores + age + elapsed time |
| `pft_fev1q()` | Adult mortality index, alternative to CCS in adults | FEV1 (litres) + sex |

For multi-point trajectory fitting (slopes, mixed-effects models,
disease-specific decline thresholds), use the right tool for the
job directly — `stats::lm()`, `lme4::lmer()`, or any of the
general-purpose longitudinal-modelling packages. Those decisions
depend on cohort design, covariates, and nesting structure in ways
that don't generalise into a one-size-fits-all PFT wrapper.

# 1. Two-point conditional change score: `pft_change()`

`pft_change()` evaluates whether the change between *two* FEV1
z-scores is larger than expected from within-subject variability and
regression to the mean. The formula (Box 2 of Stanojevic 2022):

\[
\text{CCS} = \frac{z_2 - r \cdot z_1}{\sqrt{1 - r^2}}, \quad
r = 0.642 - 0.04 \cdot \text{time}_\text{years} + 0.020 \cdot \text{age}_\text{years}.
\]

`|CCS| > 1.96` is the two-sided 95 % significance threshold:

```{r}
# Box 2 worked example: 14-year-old male, FEV1 z dropped from -0.78
# to -1.60 over 3 months.
pft_change(z1 = -0.78, z2 = -1.60, age_t1 = 14, time_years = 0.25)
```

The same drop spread over four years is not significant -- the
autocorrelation falls, so the same z-score change is more readily
explained by noise:

```{r}
pft_change(z1 = -0.78, z2 = -1.60, age_t1 = 14, time_years = 4)
```

**Scope.** The formula was derived in a pediatric / young-adult
cohort; the standard notes it has "yet to be validated, extended to
adults" but allows it as "a reasonable tool to facilitate
interpretation". For adults the standard recommends FEV1Q instead
(Box 3); see section 2 below.

# 2. Adult FEV1Q: `pft_fev1q()`

For *adults*, the 2022 standard recommends FEV1Q over the conditional
change score: the ratio of FEV1 to a sex-specific 1st-percentile
"bottom line" required for survival (Box 3). FEV1Q is not a change
score -- it characterises the *absolute* impairment relative to a
survivable lower limit, in a way that's interpretable across
patients and over time.

```{r}
# Box 3 worked example: 70-year-old female with FEV1 = 0.9 L.
pft_fev1q(fev1 = 0.9, sex = "F", age = 70)
```

Values closer to 1 indicate higher mortality risk; under normal
ageing one unit of FEV1Q is lost approximately every 18 years (10
years in smokers and the elderly). The function refuses age < 18 by
returning `NA_real_` per the paper's "not appropriate for children
and adolescents" caveat.

# 3. Plotting trajectories

`pft_plot()` itself only produces single-patient lollipop figures.
For longitudinal trajectories, pipe a long-form `pft_long()` (or
hand-built) table into `ggplot2` directly:

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
serial <- data.frame(
  patient_id  = rep(1:2, each = 4),
  visit_date  = rep(as.Date(c("2020-01-15","2021-03-10",
                               "2022-05-20","2023-07-30")), 2),
  fev1_zscore = c(-0.5, -0.8, -1.2, -1.6,
                   0.2,  0.0, -0.3, -0.5)
)
ggplot(serial, aes(visit_date, fev1_zscore,
                   colour = factor(patient_id), group = patient_id)) +
  geom_hline(yintercept = -1.645, linetype = "dotted") +
  geom_line() + geom_point() +
  labs(x = "Visit date", y = "FEV1 z-score", colour = "Patient") +
  theme_minimal()
```

# See also

* `vignette("interpretation-guide")` -- severity bands, pattern
  decision tree, 2022 vs 2005.
* `vignette("diffusion-capacity")` -- DLCO interpretation and
  Hb correction.
* `?pft_change`, `?pft_fev1q` for the function references.
