---
title: "Getting started with pft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

`pft` implements the Stanojevic et al. ERJ 2022 ERS/ATS interpretive
strategy for pulmonary function tests: reference values, z-scores,
percent predicted, ATS pattern classification, severity grading,
bronchodilator response, PRISm screening, and conditional change
scores, all from a data-frame-pipeline API.

The sections below run the pipeline on a single patient and then on
a small cohort.

```{r}
library(pft)
```

# 1. Reference values from demographics alone

The simplest call: pass age, sex, height (and race for GLI 2012) and
get predicted values, lower limits of normal (LLN), and upper limits of
normal (ULN) for every measure.

```{r}
patient <- data.frame(
  sex = "M", age = 45, height = 178, race = "Caucasian"
)

ref <- pft_spirometry(patient, year = 2012)
ref[, c("fev1_pred", "fev1_lln", "fev1_uln",
        "fvc_pred",  "fvc_lln",  "fvc_uln")]
```

The same pattern works for lung volumes and diffusion:

```{r}
pft_volumes(patient)[, c("frc_pred", "tlc_pred", "rv_pred", "vc_pred")]
pft_diffusion(patient)[, c("dlco_pred", "kco_tr_pred", "va_pred")]
```

# 2. Z-scores and percent predicted from measured values

Add `<measure>_measured` columns and z-scores and percent-predicted
appear automatically next to the reference values.

```{r}
patient_with_measurements <- data.frame(
  sex = "M", age = 45, height = 178, race = "Caucasian",
  fev1_measured = 2.5,
  fvc_measured  = 3.8
)

out <- pft_spirometry(patient_with_measurements)
out[, c("fev1_pred", "fev1_zscore", "fev1_pctpred",
        "fvc_pred",  "fvc_zscore",  "fvc_pctpred")]
```

The z-score uses the standard LMS formula
`((measured / M)^L - 1) / (L * S)`. Percent predicted is
`(measured / M) * 100`.

# 3. Severity grading

`pft_severity()` maps a z-score to one of four categories per the
Stanojevic 2022 cut points:

```{r}
pft_severity(c(0, -1.7, -3, -5))
```

You can grade any z-score column directly:

```{r}
out$fev1_severity <- pft_severity(out$fev1_zscore)
out$fvc_severity  <- pft_severity(out$fvc_zscore)
out[, c("fev1_zscore", "fev1_severity", "fvc_zscore", "fvc_severity")]
```

# 4. ATS pattern classification

Given measured spirometry plus TLC and their LLNs, `pft_classify()`
labels the pattern per Stanojevic 2022 Figure 8:

```{r}
classification_input <- data.frame(
  fev1 = 2.5,  fev1_lln = 3.0,
  fvc  = 3.8,  fvc_lln  = 3.5,
  fev1fvc = 0.66, fev1fvc_lln = 0.70,
  tlc  = 6.0,  tlc_lln = 5.0
)
pft_classify(classification_input)[
  , c("ats_classification", "ats_pattern_combination")
]
```

The 4-character `ats_pattern_combination` records which inputs drove the
label (A = abnormal / below LLN, N = at or above LLN), in the order
FEV1, FVC, FEV1/FVC, TLC. `ANAN` above means FEV1 and FEV1/FVC are low;
FVC and TLC are normal.

# 5. Bronchodilator response

The Stanojevic 2022 BDR criterion is a >10% change relative to predicted
in FEV1 or FVC (replacing the 2005 12% / 200 mL rule):

```{r}
pft_bdr(pre = 2.5, post = 3.0, predicted = 4.0)
```

# 6. PRISm screening

Preserved Ratio Impaired Spirometry: low FEV1 with normal FEV1/FVC.
Spirometry-only; no TLC needed.

```{r}
pft_prism(data.frame(
  fev1    = 2.0,  fev1_lln    = 2.5,
  fvc     = 2.6,  fvc_lln     = 3.0,
  fev1fvc = 0.80, fev1fvc_lln = 0.70
))
```

# 7. Serial change

For longitudinal monitoring, the conditional change score (CCS)
adjusts for regression to the mean using a within-subject z-score
autocorrelation `r`. `|CCS| > 1.645` indicates an unusual change.

```{r}
# z dropped from -0.5 to -2.5 over 1 year; r ≈ 0.7 for adult FEV1
pft_change(z1 = -0.5, z2 = -2.5, r = 0.7)
```

# 8. Input QC

`pft_validate()` flags biologically implausible inputs without
erroring -- handy for screening a real-world cohort before
interpretation.

```{r}
cohort <- data.frame(
  sex    = c("M", "X", "F"),
  age    = c(45, 30, 250),
  height = c(178, 30, 165),
  fev1_measured = c(2.5, 4.0, 2.0),
  fvc_measured  = c(3.8, 3.5, 2.5)
)
pft_validate(cohort)[, c("qc_pass", "qc_issues")]
```

# 9. The one-call workflow

`pft_interpret()` auto-detects every available input and produces the
full Stanojevic 2022-compliant interpretation in one call:

```{r}
patient <- data.frame(
  sex = "M", age = 45, height = 178, race = "Caucasian",
  fev1_measured     = 2.5,
  fvc_measured      = 3.8,
  fev1fvc_measured  = 2.5 / 3.8,
  tlc_measured      = 6.0,
  fev1_pre          = 2.5,
  fev1_post         = 2.9
)
result <- pft_interpret(patient)

# A clinically useful subset of the ~60 columns generated:
result[, c("fev1_pred", "fev1_zscore", "fev1_severity",
           "fvc_zscore", "fvc_severity",
           "ats_classification", "prism",
           "fev1_bdr_pct", "fev1_bdr_significant")]
```

# 10. Visualisation

`pft_plot()` produces a clinical-style z-score lollipop figure with
severity bands. Requires `ggplot2` (Suggests).

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
pft_plot(result)
```

# 11. Cohort analyses

Everything composes naturally in a pipeline. Apply `pft_interpret()`
to a multi-row data frame and the output is the same data frame with
~60 interpretation columns appended:

```{r}
cohort <- data.frame(
  sex    = c("M", "F", "M"),
  age    = c(45, 60, 30),
  height = c(178, 165, 175),
  race   = c("Caucasian", "AfrAm", "Caucasian"),
  fev1_measured = c(2.5, 1.8, 4.0),
  fvc_measured  = c(3.8, 2.4, 5.2),
  fev1fvc_measured = c(2.5/3.8, 1.8/2.4, 4.0/5.2),
  tlc_measured  = c(6.0, 4.5, 6.8)
)

interpreted <- pft_interpret(cohort)
interpreted[, c("sex", "age",
                "fev1_zscore", "fev1_severity",
                "ats_classification", "prism")]
```

# 12. Long-form tidiers for downstream analysis

`pft_long()` pivots a wide `pft_result` into one row per
`(patient, measure)`, with the natural shape for `dplyr` /
`ggplot2` workflows. `pft_glance()` returns a per-patient summary
(top-level interpretation columns plus derived stats like
`worst_zscore`, `n_below_lln`).

```{r}
pft_long(interpreted)[1:6, ]
pft_glance(interpreted)
```

The S3 methods `tidy.pft_result()` and `glance.pft_result()`
dispatch to these when `broom` is installed, so
`broom::tidy(interpreted)` works identically to `pft_long()`.

# 13. Composite pattern + severity labels

`pft_pattern_severity()` composes the ATS pattern with the
driving measure's severity into a single label like
`"Moderate Obstructed"` or `"Severe Mixed"`, useful for
clinic-note rendering. `pft_interpret()` runs it automatically;
you can also call it on its own.

```{r}
interpreted$pattern_severity
```

# 14. Diffusion clinical category

When `pft_diffusion()` outputs are available (the default in
`pft_interpret()` when demographics are supplied), the Hughes &
Pride 2012 categorical interpretation falls out of `dlco_zscore`,
`va_zscore`, `kco_*_zscore`:

```{r}
patient_dlco <- data.frame(
  sex = "M", age = 50, height = 178, race = "Caucasian",
  dlco_measured   = 6,      # low
  va_measured     = 6,
  kco_tr_measured = 1.0     # also low -> Parenchymal pattern
)
pft_interpret(patient_dlco)$diffusion_category
```

# 15. Comparing GLI 2012 vs GLI Global 2022

For institutions migrating from race-stratified GLI 2012 to
race-neutral GLI Global 2022, `pft_compare()` runs both equation
sets on the same input and emits per-row reclassification
deltas plus a `summary()` method for cohort-level audits:

```{r}
cmp <- pft_compare(cohort)
cmp[, c("sex", "race",
        "fev1_zscore", "fev1_zscore_2022", "fev1_zscore_delta",
        "ats_classification", "ats_classification_2022",
        "ats_pattern_changed")]
summary(cmp)
```

# 16. Longitudinal slope fitting

For three or more serial measurements per patient,
`pft_decline()` fits per-patient slopes (OLS by default, with
`lme4`-backed mixed-effects available) and flags patients whose
decline crosses a user-supplied threshold.

```{r}
serial <- data.frame(
  patient_id = rep(c("P1", "P2", "P3"), each = 5),
  year       = rep(2018:2022, times = 3),
  fev1_zscore = c(
    -0.5, -0.4, -0.6, -0.5, -0.5,     # P1 stable
    -0.5, -0.9, -1.3, -1.7, -2.1,     # P2 declining
     0.2, -0.65, -1.5, -2.3, -3.2     # P3 rapid decline
  )
)
pft_decline(serial, by = patient_id, measure = "fev1_zscore",
            time = year, flag_threshold = 0.25)
```

# 17. Lung age

For patient counseling, `pft_lung_age()` solves for the age that
would produce the patient's measured FEV1 as the GLI predicted
value at their height and sex. A 45-year-old whose FEV1 is what a
65-year-old's would normally be has a lung age of 65.

```{r}
counsel <- data.frame(
  sex = "M", age = 45, height = 178, race = "Caucasian",
  fev1_measured = 2.8     # low for a 45-year-old
)
pft_lung_age(counsel)[, c("age", "fev1_lung_age", "fev1_lung_age_delta")]
```

# 18. Unit auto-detection

The two most common clinic / lab data-import bugs that yield
silently wrong reference values are height-in-inches (instead of
cm) and volumes-in-millilitres (instead of L).
`pft_normalize_units()` catches both with conservative
heuristics, warns about the conversion, and returns a corrected
data frame ready for the rest of the pipeline:

```{r}
wrong_units <- data.frame(
  sex = "M", age = 45, height = 70,   # 70 inches!
  race = "Caucasian",
  fev1_measured = 2500                # millilitres!
)
suppressWarnings(pft_normalize_units(wrong_units))
```

# 19. Schema introspection

`pft_schema()` documents every output column the pipeline can
produce for a given configuration -- handy when building forms
or building integrations against the package programmatically:

```{r}
schema <- pft_schema(year = 2012, SI.units = FALSE)
head(schema, 10)
```

# 20. Cohort-level audits

`pft_cohort_summary()` summarises a `pft_result` (or compatible
data frame) cohort-wide; pass `by = "sex"` (or any other column
vector) for stratification:

```{r}
pft_cohort_summary(interpreted)
pft_cohort_summary(interpreted, by = "sex")
```

When the data contains both 2012 and 2022 classifications (e.g.,
a `pft_compare()` output), the summary gains a
`reclassification` component containing the 2012 -> 2022
confusion matrix.

# 21. Multi-mode visualisation

`pft_plot()` supports five modes:

* `"lollipop"` (default, single patient; section 10 above)
* `"histogram"` -- cohort z-score distribution per measure
* `"trajectory"` -- longitudinal z-score per measure over time
* `"bdr"` -- pre/post bronchodilator paired arrows
* `"compare"` -- GLI 2012 -> GLI Global 2022 reclassification
  arrows (consumes a `pft_compare()` result or any data with
  paired `_zscore` and `_zscore_2022` columns)

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
pft_plot(interpreted, type = "histogram")
```

# Citations

See `citation("pft")` for the package and underlying reference
standards as `bibentry` objects, suitable for direct inclusion in
publications.

```{r}
citation("pft")
```
