---
title: "Getting started with pft"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with pft}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4
)
```

`pft` is an R toolkit for **ERS/ATS 2022-compliant pulmonary function
test (PFT) interpretation**. It implements the full Stanojevic et al.
ERJ 2022 interpretive strategy: reference values, z-scores, percent
predicted, ATS pattern classification, severity grading, bronchodilator
response, PRISm screening, and conditional change scores -- all from a
data-frame-pipeline API.

This vignette walks through the full workflow on a single patient and
then on a small cohort.

```{r}
library(pft)
```

# 1. Reference values from demographics alone

The simplest call: pass age, sex, height (and race for GLI 2012) and
get predicted values, lower limits of normal (LLN), and upper limits of
normal (ULN) for every measure.

```{r}
patient <- data.frame(
  sex = "M", age = 45, height = 178, race = "Caucasian"
)

ref <- spirometry_normals(patient, year = 2012)
ref[, c("fev1_pred", "fev1_lln", "fev1_uln",
        "fvc_pred",  "fvc_lln",  "fvc_uln")]
```

The same pattern works for lung volumes and diffusion:

```{r}
volume_normals(patient)[, c("frc_pred", "tlc_pred", "rv_pred", "vc_pred")]
diffusion_normals(patient)[, c("dlco_pred", "kco_tr_pred", "va_pred")]
```

# 2. Z-scores and percent predicted from measured values

Add `<measure>_measured` columns and z-scores and percent-predicted
appear automatically next to the reference values.

```{r}
patient_with_measurements <- data.frame(
  sex = "M", age = 45, height = 178, race = "Caucasian",
  fev1_measured = 2.5,
  fvc_measured  = 3.8
)

out <- spirometry_normals(patient_with_measurements)
out[, c("fev1_pred", "fev1_zscore", "fev1_pctpred",
        "fvc_pred",  "fvc_zscore",  "fvc_pctpred")]
```

The z-score uses the standard LMS formula
`((measured / M)^L - 1) / (L * S)`. Percent predicted is
`(measured / M) * 100`.

# 3. Severity grading

`severity_grade()` maps a z-score to one of four categories per the
Stanojevic 2022 cut points:

```{r}
severity_grade(c(0, -1.7, -3, -5))
```

You can grade any z-score column directly:

```{r}
out$fev1_severity <- severity_grade(out$fev1_zscore)
out$fvc_severity  <- severity_grade(out$fvc_zscore)
out[, c("fev1_zscore", "fev1_severity", "fvc_zscore", "fvc_severity")]
```

# 4. ATS pattern classification

Given measured spirometry plus TLC and their LLNs, `ats_classification()`
labels the pattern per Stanojevic 2022 Figure 8:

```{r}
classification_input <- data.frame(
  fev1 = 2.5,  fev1_lln = 3.0,
  fvc  = 3.8,  fvc_lln  = 3.5,
  fev1fvc = 0.66, fev1fvc_lln = 0.70,
  tlc  = 6.0,  tlc_lln = 5.0
)
ats_classification(classification_input)[
  , c("ats_classification", "ats_pattern_combination")
]
```

The 4-character `ats_pattern_combination` records which inputs drove the
label (A = abnormal / below LLN, N = at or above LLN), in the order
FEV1, FVC, FEV1/FVC, TLC. `ANAN` above means FEV1 and FEV1/FVC are low;
FVC and TLC are normal.

# 5. Bronchodilator response

The Stanojevic 2022 BDR criterion is a >10% change relative to predicted
in FEV1 or FVC (replacing the 2005 12% / 200 mL rule):

```{r}
bronchodilator_response(pre = 2.5, post = 3.0, predicted = 4.0)
```

# 6. PRISm screening

Preserved Ratio Impaired Spirometry: low FEV1 with normal FEV1/FVC.
Spirometry-only; no TLC needed.

```{r}
prism_screen(data.frame(
  fev1 = 2.0,    fev1_lln = 2.5,
  fev1fvc = 0.80, fev1fvc_lln = 0.70
))
```

# 7. Serial change

For longitudinal monitoring, the conditional change score (CCS)
adjusts for regression to the mean using a within-subject z-score
autocorrelation `r`. `|CCS| > 1.645` indicates an unusual change.

```{r}
# z dropped from -0.5 to -2.5 over 1 year; r ≈ 0.7 for adult FEV1
serial_change_score(z1 = -0.5, z2 = -2.5, r = 0.7)
```

# 8. Input QC

`validate_pft()` flags biologically implausible inputs without
erroring -- handy for screening a real-world cohort before
interpretation.

```{r}
cohort <- data.frame(
  sex    = c("M", "X", "F"),
  age    = c(45, 30, 250),
  height = c(178, 30, 165),
  fev1_measured = c(2.5, 4.0, 2.0),
  fvc_measured  = c(3.8, 3.5, 2.5)
)
validate_pft(cohort)[, c("qc_pass", "qc_issues")]
```

# 9. The one-call workflow

`pft_interpret()` auto-detects every available input and produces the
full Stanojevic 2022-compliant interpretation in one call:

```{r}
patient <- data.frame(
  sex = "M", age = 45, height = 178, race = "Caucasian",
  fev1_measured     = 2.5,
  fvc_measured      = 3.8,
  fev1fvc_measured  = 2.5 / 3.8,
  tlc_measured      = 6.0,
  fev1_pre          = 2.5,
  fev1_post         = 2.9
)
result <- pft_interpret(patient)

# A clinically useful subset of the ~60 columns generated:
result[, c("fev1_pred", "fev1_zscore", "fev1_severity",
           "fvc_zscore", "fvc_severity",
           "ats_classification", "prism",
           "fev1_bdr_pct", "fev1_bdr_significant")]
```

# 10. Visualisation

`plot_pft()` produces a clinical-style z-score lollipop figure with
severity bands. Requires `ggplot2` (Suggests).

```{r, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_pft(result)
```

# 11. Cohort analyses

Everything composes naturally in a pipeline. Apply `pft_interpret()`
to a multi-row data frame and the output is the same data frame with
~60 interpretation columns appended:

```{r}
cohort <- data.frame(
  sex    = c("M", "F", "M"),
  age    = c(45, 60, 30),
  height = c(178, 165, 175),
  race   = c("Caucasian", "AfrAm", "Caucasian"),
  fev1_measured = c(2.5, 1.8, 4.0),
  fvc_measured  = c(3.8, 2.4, 5.2),
  fev1fvc_measured = c(2.5/3.8, 1.8/2.4, 4.0/5.2),
  tlc_measured  = c(6.0, 4.5, 6.8)
)

interpreted <- pft_interpret(cohort)
interpreted[, c("sex", "age",
                "fev1_zscore", "fev1_severity",
                "ats_classification", "prism")]
```

# Citations

See `citation("pft")` for the package and underlying reference
standards as `bibentry` objects, suitable for direct inclusion in
publications.

```{r}
citation("pft")
```
